---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
title: "Assignment 1: Online News Popularity"
author:
- name: Khoi Mai Tieu
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
endnote: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      message=FALSE, warning=FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })
```


# Introduction

In this assignement, I choose to work with the [Online News Popularity Dataset](https://archive.ics.uci.edu/dataset/332/online+news+popularity)
a dataset consisting of 39,000 articles sourced from the Mashable website, covering a two-year period from January 7, 2013, to January 7, 2015
This dataset summarizes a heterogeneous set of features about articles published by Mashable in a period of two years. The goal is to predict the number of shares in social networks (popularity).

The data was collected during a two year period, from January 7 2013 to January 7 2015.


```{r}
download.file('http://archive.ics.uci.edu/ml/machine-learning-databases/00332/OnlineNewsPopularity.zip', 'OnlineNewsPopularity.zip')

data <- read.csv(unz('OnlineNewsPopularity.zip',
                     'OnlineNewsPopularity/OnlineNewsPopularity.csv'),
                 header = TRUE)
system('rm OnlineNewsPopularity.zip')
```



# Problem 1. 
```{r}
library(vtable)
st(data)
```




TBD