---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
title: "Assignment 1: Online News Popularity"
author:
- name: Khoi Mai Tieu
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
endnote: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      message=FALSE, warning=FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })
```


# Introduction

In this assignement, I choose the [Online News Popularity Dataset](https://archive.ics.uci.edu/dataset/332/online+news+popularity) consisting of 39,000 articles sourced from the Mashable website, covering a two-year period from January 7, 2013, to January 7, 2015.
This dataset summarizes a heterogeneous set of features about articles published by Mashable in a period of two years.

```{r}
download.file('http://archive.ics.uci.edu/ml/machine-learning-databases/00332/OnlineNewsPopularity.zip', 'OnlineNewsPopularity.zip')

data <- read.csv(unz('OnlineNewsPopularity.zip',
                     'OnlineNewsPopularity/OnlineNewsPopularity.csv'),
                 header = TRUE)
system('rm OnlineNewsPopularity.zip')
```

# Problem 1. Summary Statistics Table
```{r}
library(vtable)
st(data)
```

There are 61 attributes including

- 2 non-predictive (url, timedelta)
- 58 predictors which is categorized into 6 groups
  - Words: 5 predictors (n_tokens_title, n_tokens_content, n_unique_tokens, avg_token_length, n_non_stop_words, n_non_stop_unique_tokens)
  - Links: 5 predictors (num_hrefs, num_self_hrefs, self_reference_min/max/avg_shares)
  - Digital Media: 2 predictors (num_imgs, num_videos)
  - Time: 8 predictors (weekday_is_XXX, is_weekend)
  - Keywords: 16 predictors (data_channel_is_XXX, num_keywords, kw_min/max/avg_min, kw_min/max/avg_max, kw_min/max/avg_avg)
  - Natural Language Processing: 22 predictors
- 1 goal field (shares)

| Category                       | Feature                                              | Type (#)  |
|--------------------------------|------------------------------------------------------|-----------|
| **Non-predictive**             | URL of the article                                   | text (1)  |
|                                | Days between the article publication and the dataset acquisition | time (1)|
| **Words**                      | Number of words in the title                         | number (1)|
|                                | Number of words in the article                       | number (1)|
|                                | Average word length                                  | number (1)|
|                                | Rate of non-stop words                               | ratio (1) |
|                                | Rate of unique words                                 | ratio (1) |
|                                | Rate of unique non-stop words                        | ratio (1) |
| **Links**                      | Number of links                                      | number (1)|
|                                | Number of Mashable article links                     | number (1)|
|                                | Minimum, average and maximum number of shares of Mashable links | number (3) |
| **Digital Media**             | Number of images                                     | number (1)|
|                                | Number of videos                                     | number (1)|
| **Time**                      | Day of the week                                      | nominal (1)|
|                                | Published on a weekend?                              | bool (1)  |
| **Keywords**                   | Number of keywords                                   | number (1)|
|                                | Worst keyword (min./avg./max. shares)               | number (3)|
|                                | Average keyword (min./avg./max. shares)             | number (3)|
|                                | Best keyword (min./avg./max. shares)                | number (3)|
|                                | Article category (lifestyle/entertainment/bus/socmed/tech/world) | nominal (1)|
| **Natural Language Processing** | Closeness to top 5 LDA topics                        | ratio (5) |
|                                | Title subjectivity                                   | ratio (1) |
|                                | Article text subjectivity score and its absolute difference to 0.5 | ratio (2) |
|                                | Title sentiment polarity                             | ratio (1) |
|                                | Rate of positive and negative words                  | ratio (2) |
|                                | Pos. words rate among non-neutral words              | ratio (1) |
|                                | Neg. words rate among non-neutral words              | ratio (1) |
|                                | Polarity of positive words (min./avg./max.)         | ratio (3) |
|                                | Polarity of negative words (min./avg./max.)         | ratio (3) |
|                                | Article text polarity score and its absolute difference to 0.5 | ratio (2) |
| **Target**                     | Number of article Mashable shares                    | number (1)|

The goal is to predict the number of shares in social networks (popularity). 

# Problem 2. Bad Data Visualization

TBD

# Problem 3. Good Data Visualization

TBD

# Problem 4. Simple analysis

TBD